---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# graphonmix

 <!-- badges: start -->
  [![R-CMD-check](https://github.com/sevvandi/graphonmix/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sevvandi/graphonmix/actions/workflows/R-CMD-check.yaml)
  <!-- badges: end -->

The goal of graphonmix is to generate $(U,W)$-mixture graphs. These graphs are generated from a standard dense graphon $W$ and a disjoint clique graphons $U$, which acts as the sparse graphon. 

## Installation

You can install the development version of graphonmix from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("sevvandi/graphonmix")
```

## Example

This is a basic example on how to sample a $(U,W)$-mixture graph.

```{r example}
library(graphonmix)
library(igraph)
# create the dense graphon W(x,y) = 0.1
W <- matrix(0.1, nrow = 100, ncol = 100)
# create the sparse part - a disjoint set of stars
wts <- c(0.5, 0.3, 0.2)
# single function to generate a graph mixture
gr1 <- sample_mixed_graph(W, wts, nd = 100, ns = 300, p = 0.5, option = 2)
plot(gr1,
     edge.curved = 0.3,
     vertex.size = degree(gr1)*0.1,
     edge.color = "lightgray",     # Light colored edges
     vertex.label = NA,
     vertex.color = "lightblue",
     main = "(U,W) Graph mixture"
)
```

Or you can generate the two graphs separately and join them. 

```{r}
# sample the dense part and plot
grdense <- sample_graphon(W, 100)
plot(grdense,
     edge.curved = 0.3,
     vertex.size = degree(grdense)*0.1,
     edge.color = "lightgray",     # Light colored edges
     vertex.label = NA,
     vertex.color = "lightblue",
     main = "Dense Part"
)

# sample the sparse part and plot
grsparse <- generate_star_union(wts, 300)
plot(grsparse,
     edge.curved = 0.3,
     vertex.size = degree(grsparse)*0.1,
     edge.color = "lightgray",     # Light colored edges
     vertex.label = NA,
     vertex.color = "lightblue",
     main = "Sparse Part"
)

# join the two graphs and plot
gr2 <- graph_join(grdense, grsparse, option = 2)
plot(gr2,
     edge.curved = 0.3,
     vertex.size = degree(gr2)*0.1,
     edge.color = "lightgray",     # Light colored edges
     vertex.label = NA,
     vertex.color = "lightblue",
     main = "(U,W) Graph mixture"
)
```

